---
title: "appendix_section"
author: "Danny Luo"
date: "4/30/2021"
header-includes: 
  - \def\bs{\boldsymbol}
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```







## Hyper Prior elicitation for V for package mixAK

MixAK only allows a uniform parameter $V$, and gives the option to specify $V$ only through specifying the hyperprior parameters of a Gamma distribution.[@R-mixAK] According to the documentation page of the package, matrix $V$ is assumed to be diagonal with $\gamma_1,\gamma_2,...,\gamma_p$  on the diagonal, and for each $\gamma_j,$ $\gamma_j^{-1} \sim Gamma(g_j,h_j)$.

We checked that the covariance matrix of some ramdom subset of the population are approximatley diagonal. Since matrix $V$ is the scale matrix oarameters of the Wishart prior placed on cluster covariance matrix $\Sigma_j$ and that it does not change with different groups, we performed the following elicitation:.

- Draw 1000 random samples of size 1000 from the dataset and calulated the 5 by 5 covariance matrix $C_i$ for $i =1,2,...1000$.
- For j =1,2,..,5, extract samples $L_j=\{C_i[j,j]\}^{1000}_{i=1}$.
- Forj =1,2,..,5, fit a gamma distribution using $L_j$, using the r package $fitdistrplus$. [@R-fitdistrplus] [@R-fitdistrplus-article] and using the estimation from the output to be our $\{ g_j,h_j \}$

The fitting was highly accurate as we checked the Q-Q plot to be almost matching.


## MCMC Diagnostics for cluster means and weights

```{r}
#read result 
post_data <- readRDS("Posterior_Final.rds")
```

```{r fig.height=10,fig.width=6,out.width = "70%",fig.align="center" ,fig.cap="Traceplots and Autocorrelation plots for cluster means"}
##MCMC Diagnostics for means
N=5000


col_names=colnames(post_data$mu)


par(mfrow=c(5,2),mar=c(5,5,5,5))

for(i in 1:length(col_names)){
  plot(1:N,post_data$mu[,i],type="l",ylab=col_names[i],xlab="Iter")
  acf(post_data$mu[,i],ylab=col_names[i],main="")
}

```


```{r fig.height=10,fig.width=6,out.width = "70%",fig.cap="Traceplots and Autocorrelation plots for weights",fig.align="center"}
##MCMC Diagnostics for weights
library(coda)
N=5000


col_names=colnames(post_data$w)

par(mfrow=c(5,2),mar=c(5,5,5,5))

for(i in 1:length(col_names)){
  plot(1:N,post_data$w[,i],type="l",ylab=col_names[i],xlab="Iter")
  acf(post_data$w[,i],ylab=col_names[i],main="")
}
```













